---
title: "DESeq Analysis"
author: "Alessandro Desy"
date: "`r Sys.Date()`"
output:
   html_document:
    toc: true
    toc_depth: 2
    theme: united
    number_section : no
    toc_float:
      collapsed: no
      smooth_scroll : yes
bibliography: ../Bibliography/biblio.bib  
csl: ../Bibliography/apa.csl 
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}

#the following line is to be changed based on your setup
your_folder="H:/DRT/PHYSIQUE ET GÉNIE BIOMÉDICAL/10.UTILISATEURS/Alessandro/Etudiants/Bioinformatique_Marine"
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = your_folder) 
#notes on duplicates:
#https://help.galaxyproject.org/t/too-many-duplicated-sequences-or-unassigned-because-of-mapping-quality/11965/3
#https://dnatech.ucdavis.edu/faqs/should-i-remove-pcr-duplicates-from-my-rna-seq-data#:~:text=For%20example%20plant%20RNA%2Dseq,%2Dcell%20RNA%2Dseq%20protocols.
#https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-018-4933-1#:~:text=Unique%20molecular%20identifiers%20(UMIs)%20are,need%20to%20have%20different%20UMIs.
```

# Introduction

This document presents the different DESeq analysis that were performed in the work of Merlin M. et al.

This document was based on the informations provided by the following : 

- Computational Genomics Course of Université Laval
- Bioconductor Helper : https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

## Setup

We download the different libraries of use and set up the environment.
```{r packages_import, echo= T, results = 'hide', message=FALSE, warning=FALSE}
library(DESeq2)
library(readxl)
library(tidyverse)
library(ggplot2)
library(ComplexHeatmap)
library(apeglm)
library(pheatmap)
library("RColorBrewer")
library(circlize)
```

# Raw Data
We import the raw data for this study.

```{r raw_data_import}
#to adapt to your pathway :
raw_file="./Data/raw_counts.xlsx"

raw_counts <- read_excel(raw_file) 
print(paste0("The raw data number of lines is : ",nrow(raw_counts)))
#we only keep non duplicate values
raw_counts <- distinct(raw_counts)
print(paste("After removing the duplicate lines, the number of lines drops to : ",nrow(raw_counts)))
raw_counts_mat<- raw_counts[, -1] %>% as.matrix
rownames(raw_counts_mat)<- raw_counts$symbol
head(raw_counts_mat)
```
In this file, the first two columns are the control group ; and the two latest are the TMZ (temozolomid treatred mice) group to be compared. We create the coldata data frame with this information and we give its row names the same values as the column names of raw_counts_mat. 

```{r coldata_setup}
coldata<- data.frame(condition = c("Control", "Control", "TMZ", "TMZ"))
 
rownames(coldata)<- colnames(raw_counts_mat)

head(coldata) 
print(paste("Is the coldata rownames order similar to raw_counts_mat colnames order ?", all(rownames(coldata) == colnames(raw_counts_mat)))) 
```
# DESeq object

## DESeqDataSet creation

Now we can run the DESeq2 script [@love2014moderated].

```{r deseqdataset_run}

dds <- DESeqDataSetFromMatrix(countData = raw_counts_mat,
                              colData = coldata,
                              design = ~ condition) 
print(dds) 
```

Based on the following, https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering, we prefilter the dds results by keeping only rows where at minimum 10 counts are measured in at least 3 groups. 
```{r deseq_pre_filter}
keep <- rowSums(counts(dds)>= 10) >=3 
print(paste("With this filter,",sum(keep==T),"genes are kept (",round(sum(keep==T)*100/length(keep),2),"%)"))
dds <- dds[keep,]
```
The default reference level is chosen alphabetically (i.e, in this case, "Control"). But just to ensure this, we run the following line.

```{r level_set_up}
print(paste('In our dds, ',levels(dds$condition)[1],' is the reference and',
            levels(dds$condition)[2], ' is compared to it.'))
```
## DESeq run
Now we can run the differential expression analysis.

```{r deseq_run, message=FALSE, warning=FALSE}
dds <- DESeq(dds) 
```

The dds results are saved in the following res_xxx variables, with the default results method saved in the res variable. By default, independent filtering is performed to select a set of genes for multiple test correction which maximizes the number of adjusted p-values less than a given critical value alpha (by default 0.1)

```{r deseq_results}
res <- results(dds, contrast = c("condition", "TMZ", "Control")) 
res_BY <- results(dds, contrast = c("condition", "TMZ", "Control"),pAdjustMethod='BY') 
res_BH <- results(dds, contrast = c("condition", "TMZ", "Control"),pAdjustMethod='BH') 
res_fdr <- results(dds, contrast = c("condition", "TMZ", "Control"),pAdjustMethod='fdr') 
res_bonferroni <- results(dds, contrast = c("condition", "TMZ", "Control"),pAdjustMethod='bonferroni')
res_holm <- results(dds, contrast = c("condition", "TMZ", "Control"),pAdjustMethod= "holm") 
```

In the previous lines, we defined each res_xxx variable based on different adjusted p-values, with the default being BH. These results files have the following structure :
```{r res_xxx_structure}
res

write.csv(as.data.frame(res),"./Output_R_Aless/dds_res_unique.csv", row.names = T)
```

- **baseMean** is the average of normalized counts ;

- **log2FoldChange** is positive when the gene is upregulated in the TMZ group ;

- **stat** is the Wald test value for a given gene ;

- **pvalue** is the p-value of the test for a given gene (duh) ;

- **padj** is the p-value corrected for multiple-testing (?). It allows to remove false positive results based on the given method (Benjamini, Hochberg, etc.).

Based on the following : https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#alternative-shrinkage-estimators , we also define a resLFC variable. It is based on an alternative shrinkage estimator and allows for better ranking and visualisation [@zhu2019heavy].

```{r resLFC_creation}
resLFC <- lfcShrink(dds, coef="condition_TMZ_vs_Control", type="apeglm") 
```


# DESeq results analysis

## MA-plot

The following line gives a summary of the results obtained in the default res variable. By the default, the DESeq function applies an adjusted p-value cutoff of alpha < 0.1.
```{r deseq_res_summary}
summary(res)
```

And here are the results obtained in the shrunk effect size (LFC) results :

```{r deseq_resLFC_summary}
summary(resLFC)
```

As we can see, in this case it basically give the same results. But now we plot their respective MA-plot that shows the log fold change Vs the mean of normalized counts (https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#exploring-and-exporting-results).

```{r plot_ma_res}
plotMA(res, ylim=c(-2,2), main="MA-plot of res")
plotMA(resLFC, ylim=c(-2,2), main="MA-plot of shrunken log2 fold change results")
```

In the shrunken MA-plot, we had removed the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

## Volcano plot

Here are presented  the volcano plots corresponding to each DESeq result. 

```{r volcano_plots}
volcanofunc <- function(ds_res,subtt){
  ds_res$diffexpressed_padj <- "Not significative"
  ds_res$diffexpressed_padj[ds_res$log2FoldChange > 0 & ds_res$padj < 0.05] <- "Up"
  ds_res$diffexpressed_padj[ds_res$log2FoldChange < -0 & ds_res$padj < 0.05] <- "Down"
  nb_up=nrow(ds_res[ds_res$diffexpressed_padj=='Up',])
  nb_dw=nrow(ds_res[ds_res$diffexpressed_padj=='Down',])
   
  print(paste('In the',subtt, 'result file, there are', nb_up, 'genes up,',
              round(100*nb_up/nrow(ds_res),2),'%'))
  print(paste('In the',subtt, 'result file, there are', nb_dw, 'genes Down,',
              round(100*nb_dw/nrow(ds_res),2),'%'))

  # res_sig <- ds_res%>%
  #   as.data.frame()%>%
  #   tibble::rownames_to_column(var='gene')%>%
  #   filter(!stringr::str_detect(gene,'LOC'))%>%
  #   filter(abs(log2FoldChange)>=4,padj<=0.01)
    
  ggplot(data=ds_res, aes(x=log2FoldChange, y=-log10(padj),col=diffexpressed_padj)) + 
    geom_point() + 
    theme_classic() +
    theme(axis.text=element_text(size=12), legend.text = element_text(size=13),legend.position="top",legend.justification='left')+
    # geom_label(data=res_sig,aes(label=gene))+
    scale_color_manual(values=c("#4285f4", "grey", "#ea4335"))+
    scale_x_continuous(breaks = seq(-5, 5, by = 1))+
    theme(legend.title=element_blank())+
    ggtitle(paste("Volcano Plot -",subtt))
  
}

##volcano plot de BH
volcanofunc(res,'Default (BH)')
```


```{r volcano_plots_BH}
volcanofunc(res_BH, 'BH') #BH cis similar to the default one
```


```{r volcano_plots_fdr}
volcanofunc(res_fdr, 'fdr')
```


```{r volcano_plots_by}
volcanofunc(res_BY, 'BY')
```


```{r volcano_plots_bonfe}
volcanofunc(res_bonferroni, 'bonferroni')
```


```{r volcano_plots_holm}
volcanofunc(res_holm, 'holm') 
```

## Information on results

The following is a paste of https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html


For a particular gene, a log2 fold change of -1 for condition treated vs untreated means that the treatment induces a multiplicative change in observed gene expression level of $2^{−1}=0.5$
 compared to the untreated condition. If the variable of interest is continuous-valued, then the reported log2 fold change is per unit of change of that variable.


**Note on p-values set to NA**: some values in the results table can be set to NA for one of the following reasons:

- If within a row, all samples have zero counts, the baseMean column will be zero, and the log2 fold change estimates, p value and adjusted p value will all be set to NA.

- If a row contains a sample with an extreme count outlier then the p value and adjusted p value will be set to NA. These outlier counts are detected by Cook’s distance.

- If a row is filtered by automatic independent filtering, for having a low mean normalized count, then only the adjusted p value will be set to NA. 

## Variance Stabilizing Transformation and PCA

This transformation removes the dependance of the variance on the mean. We can then plot the PCA (principal components analysis). 

 

```{r vsd1}
vsd <- vst(dds, blind=FALSE) #Extracting transformed values
# head(assay(vsd), 4)

# plotPCA(vsd, intgroup=c("condition")) # by default, this only considers the first n=500 genes
plotPCA(vsd, intgroup=c("condition"),ntop=length(assay(vsd)) ) # this line allows us to create the same one as below.
```

This plot was generated by using all of the genes (n=`r length(assay(vsd))`). The source code of this PCA plot function was modfied to personnalize the colors, etc. as shown below.


```{r my_pca_plot,echo=T}
my_pca_plot_func = function(object, intgroup="condition",
                                  ntop=500, returnData=FALSE, pcsToUse=1:2)
{
  message(paste0("using ntop=",ntop," top features by variance"))
  
  # calculate the variance for each gene
  rv <- rowVars(assay(object))

  # select the ntop genes by variance
  select <- order(rv, decreasing=TRUE)[seq_len(min(ntop, length(rv)))]

  # perform a PCA on the data in assay(x) for the selected genes
  pca <- prcomp(t(assay(object)[select,]))

  # the contribution to the total variance for each component
  percentVar <- pca$sdev^2 / sum( pca$sdev^2 )

  if (!all(intgroup %in% names(colData(object)))) {
    stop("the argument 'intgroup' should specify columns of colData(dds)")
  }

  # add the intgroup factors together to create a new grouping factor 
  group <- if (length(intgroup) > 1) {
    intgroup.df <- as.data.frame(colData(object)[, intgroup, drop=FALSE])
    factor(apply( intgroup.df, 1, paste, collapse=":"))
  } else {
    colData(object)[[intgroup]]
  }

  # assembly the data for the plot
  pcs <- paste0("PC", pcsToUse)
  d <- data.frame(V1=pca$x[,pcsToUse[1]],
                  V2=pca$x[,pcsToUse[2]],
                  group=group, name=colnames(object), colData(object))
  colnames(d)[1:2] <- pcs
  
  if (returnData) {
    attr(d, "percentVar") <- percentVar[pcsToUse]
    return(d)
  }

  ggplot( data=d,aes( x = .data[[ pcs[1] ]],y = .data[[ pcs[2] ]],color=group)) +
    scale_color_manual(values=c("#45B39D","#FFC300")) + 
    geom_point(size=3) + 
    xlab(paste0(pcs[1],": ",round(percentVar[pcsToUse[1]] * 100),"% variance")) +
    ylab(paste0(pcs[2],": ",round(percentVar[pcsToUse[2]] * 100),"% variance")) +
    theme_classic() +
    theme(axis.text=element_text(size=12), legend.text = element_text(size=13),text = element_text(size = 13) )+
    geom_hline(yintercept=0)+
    geom_vline(xintercept=0)+
    coord_fixed()
}

my_pca_plot_func(vsd, intgroup=c("condition"),ntop=length(assay(vsd)) )
```
 



## Heatmap
In this section we create the default heatmap. We first select the first 20 genes in terms of average measured counts.

```{r pheatmap}
select <- order(rowMeans(counts(dds,normalized=TRUE)),decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,"condition"])

colnames(df)=c('condition')
rownames(df) <- colnames(assay(vsd))

#the following information is illustrated in the heatmap here below:
assay(vsd)[select,]
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=T,
         cluster_cols=FALSE, annotation_col=df)

```

This time we create a heatmap with genes having the highest difference of average counts when comparing Control with TMZ

```{r pheatmap2}

order_diff <- rownames(as.data.frame(counts(dds))%>%mutate(mean_diff=abs((ACt8+BCt8)/2-(ATTT8+BTTT8)/2))%>%arrange(-mean_diff))
#as.data.frame(counts(dds))%>%mutate(mean_diff=abs((ACt8+BCt8)/2-(ATTT8+BTTT8)/2)) %>%arrange(-mean_diff)


select=order_diff[0:20] #we select the first 20 genes
assay(vsd)[select,]
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=T,
         cluster_cols=FALSE, annotation_col=df)

```


## Sample-to-sample distances heatmap

A heatmap of this distance matrix gives us an overview over similarities and dissimilarities between samples. But unsure about its usefulness. 


```{r sample_dist_HM}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)


```

## Statistics

In the following, we generate a data frame that contains the average values per group (Control vs TMZ) and their absolute and relative differences.
```{r df_stats_gen}
mydf <- as.data.frame(assay(vsd)) 
mydf <- mydf%>%mutate(Ctrl_mean=(ACt8+BCt8)/2,Treat_mean=(ATTT8+BTTT8)/2)%>%
  mutate(diff_abs=Treat_mean-Ctrl_mean)%>%
  mutate(diff_rel=100*diff_abs/Ctrl_mean)
head(mydf)
write.csv(mydf,"./Output_R_Aless/mean_counts_diff_unique.csv", row.names = T)
```


The genes with the the most important positive and negative absolute differences are the following :
```{r highest_diff_abs_stat}
mydf%>%arrange(-diff_abs)%>%select(Ctrl_mean,Treat_mean,diff_abs,diff_rel)%>%head(30)
mydf%>%arrange(diff_abs)%>%select(Ctrl_mean,Treat_mean,diff_abs,diff_rel)%>%head(30)
```


The genes with the the most important positive and negative relative differences are the following :
```{r highest_diff_rel_stat}
mydf%>%arrange(-diff_rel)%>%select(Ctrl_mean,Treat_mean,diff_abs,diff_rel)%>%head(30)
mydf%>%arrange(diff_rel)%>%select(Ctrl_mean,Treat_mean,diff_abs,diff_rel)%>%head(30)
```


## Other PCA and heatmap

The following is inspired by https://www.youtube.com/watch?v=LmZD_a8XZEQ.

First, we recreate the PCA plot.
```{r pca_bis}

normalized_counts<- assay(vsd) %>%  as.matrix()
 
pca_prcomp<- prcomp(t(normalized_counts), center = TRUE, scale. = FALSE)
# str(pca_prcomp)
# pca_prcomp$x
 
PC1_and_PC2<- data.frame(PC1=pca_prcomp$x[,1], PC2= pca_prcomp$x[,2],  type = rownames(pca_prcomp$x))
ggplot(PC1_and_PC2, aes(x=PC1, y=PC2, col=type)) + 
       geom_point() + 
       geom_text(aes(label = type), hjust=0, vjust=0) +
       coord_fixed()

```

We observe that the previous graph is similar to the automatically generated plotPCA function. We can now focus on the heatmap First we only keep the significant genes (adjusted p-value <= 5% and log2FoldChange >1 or <-1).

```{r signifcant_genes}
significant_genes<- res %>% as.data.frame() %>%
  filter(padj <=0.05, abs(log2FoldChange) >= 1) %>% 
  rownames()

significant_mat<- normalized_counts[significant_genes, ] 

print(paste("When considering the signifcant genes, the number of genes went from", nrow(normalized_counts),
            "to", nrow(significant_mat), "(", round(100*nrow(significant_mat)/nrow(normalized_counts),2), "%)"))

```
```{r heat_map_bis}
col_fun = colorRamp2(c(-2, 0, 2), c("#4285F4", "white", "#EA4335"))
col_fun_8_14 = colorRamp2(c(8, 10, 14), c("#4285F4", "white", "#EA4335"))
col_anno <- HeatmapAnnotation(df = coldata, col = list(   condition= c("TMZ" = "#FFC300", "Control" = "#45B39D")))

Heatmap(t(scale(t(significant_mat))), 
        top_annotation = col_anno,
        show_row_names = FALSE, name = "scaled normalized\nexpression",
        col=col_fun )
# Heatmap(t(scale(t(significant_mat))), 
#         top_annotation = col_anno,
#         show_row_names = FALSE, name = "scaled normalized\nexpression" )

# Heatmap(t((t(significant_mat))), 
#         top_annotation = col_anno,
#         show_row_names = FALSE, name = "unscaled",
#         col=col_fun_8_14)
# Heatmap(t((t(significant_mat))), 
#         top_annotation = col_anno,
#         show_row_names = FALSE, name = "unscaled")
```


# References


 